Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-11-19 bbbq_article.tex
Changes @fransbianchi, thanks Frans

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
The peptide-binding grooves of the alleles of most haplotypes
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
therefore peptide fragments derived from human transmembrane helices (TMHs) are predicted to be presented more often as would be expected 
based on their abundance in the proteome. However, the physiological reason of why membrane proteins might be over-presented is unclear. 
In this study, we show that the predicted over-presentation of TMH-derived peptides is general, as it is predicted for bacteria and viruses 
and for both MHC-I and MHC-II, and confirmed from re-analysis of epitope databases. Moreover, we show that TMHs are evolutionarily more conserved, 
because single nucleotide polymorphisms (SNPs) are present 
relatively less frequently in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytoplasmic protein regions. 
Thus, our findings suggest that 
both cytolytic and helper T cells are more tuned to respond to membrane proteins, 
because these are evolutionary more conserved. 
We speculate that TMHs are less prone to mutations 
that enable pathogens to evade T cell responses.

@@

Our immune system fights diseases and infections from pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more specific recognition of pathogens 
than the innate immune response, 
is T cells which recognize peptides, called epitopes, derived from antigenic proteins presented on Major Histocompatibility Complexes (MHC) class I and II on the cell surface.

@@

TMHs are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of, on average, 23 amino acids in length. 
TMHs can also be predicted with high accuracy from a protein sequence 
by bioinformatics approaches \cite{krogh2001predicting,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}. 
For example, a study by Jones \cite{jones2007improving} found that,
from 184 transmembrane proteins (TMPs) with known \emph{in vitro} topology, 
80\% of the TMH predictions of these proteins
matched the experimental findings.
TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate that 15-39\% of all proteins 
in the human proteome contain at least one TMH \cite{ahram2006estimation}.
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble (i.e., extracellular or cytoplasmic) protein regions is unknown. 
In this study, we hypothesized that the presentation of 
TMH residues is evolutionarily preferred, 
since TMHs are less prone to undergo escape mutations. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their variability 
by the functional requirement to span a lipid bilayer. 
This limits 
many of the amino acids present in TMHs 
to have hydrophobic side chains 
\cite{hessa2007molecular,jones1994model}.
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as that will result in a dysfunctional TMH 
and render the protein inactive.

This study had two objectives. 
First, we aimed to generalize our findings by predicting
the antigenic presentation
from different kingdoms of life in both MHC-I and -II. 
From these \emph{in silico} predictions, we conclude that TMH-derived
epitopes from a human, viral and bacterial proteome
are likely to be presented more often than expected by chance 
for most alleles of MHC-I and II. 
We confirmed the presentation of TMH-derived peptides 
by re-analysis of peptide elution studies. 
Second, we tested our hypothesis that TMHs 
are more evolutionary conserved than soluble protein regions.
Our analysis of human single nucleotide polymorphisms (SNPs) showed 
that random point mutations are indeed less likely
to occur within TMHs. 
These findings strengthen the emerging notion 
that TMHs are important for the T cell-mediated adaptive immune system, 
and hence are of  
overlooked importance in vaccine development.

@@

To obtain experimental confirmation that peptides stemming from TMHs 
are presented \emph{in vivo} in MHC-I and MHC-II,
we mined the Immune Epitope Database (IEDB) \cite{vita2019immune}
for confirmed human MHC-ligands.
We queried the IEDB for all linear epitopes obtained
from MHC ligand assays in healthy humans, 
carrying the MHC alleles as used in this study.
From these epitopes, we kept those that were present
exactly once in the human reference proteome
with UniProt ID UP000005640\_9606.
We predicted the topology of the protein each epitope
was found in, using \verb;TMHMM; \cite{krogh2001predicting},
from which we concluded if the epitope is overlapping with a TMH
with at least 1 amino acid.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used: 
the \emph{dbSNP} \cite{sherry2001dbsnp} database,
which contains 650 million 
cataloged non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}), and the databases \emph{gene} (for gene names \cite{brown2015gene})
and \emph{protein} (for proteins sequences \cite{sayers2010database}).

@@

The Immune Epitope Database (IEDB) from the National Institutes of Health contains millions of linear epitope sequences obtained
by an MHC ligand assay.
For the MHC alleles used in this study, 
we obtained 54,303 and 2,484 linear epitope sequences for the MHC-I
and MHC-II alleles from human origin respectively.
There are relatively few epitopes for MHC-II, 
as MHC-II has many more different alleles than MHC-I,
whereas we selected only the human epitopes found for the 21 MHC-II alleles used in this study.

Figure \ref{fig:2a} shows that there are similar levels of
over-presentation of TMH-derived epitopes between (1) the
percentage of TMH-derived epitopes that is reported in the IEDB database
versus (2) the percentage of TMH-derived epitopes that is predicted to be presented
in MHC-I alleles. For MHC-II alleles, there were too few epitopes per MHC allele
to result in an informative figure. 
We showed that there is a correlation between epitopes taken from
predicted 9-mers (for MHC-I) and 14-mers (for MHC-II) from a human
reference proteome and epitopes taken from the IEDB for MHC ligands
essays (see in Supplementary figure \ref{fig:thm_presentation_correlation}).

In figure \ref{fig:2b} we grouped all the epitopes 
presented by MHC-I and MHC-II alleles 
by the percentage of TMH-derived epitopes,
which are 22\% and 10\%.

@@

\subsection{Mechanism of MHC presentation of TMH-derived epitopes}
% \paragraph{TMH presentation in vivo}

@@

  \caption{
    \textbf{
      Robust prediction that TMH epitopes are presented \emph{in vivo}.
    }
    The percentage of epitopes for MHC-I and -II alleles
    that overlap with TMHs that are presented.
    The pair of horizontal red lines in each plot indicate the lower and upper bound 
    of the 99\% confidence interval. Note that only one line is visible as this
    interval is relatively narrow.
    Alleles are listed in Table \ref{tab:haplotype_abbreviations}).
    \textbf{(A)} 
    Observed and predicted percentage of TMH-derived epitopes
    \textbf{(B)} 
    MHC ligands from IEDB binding to TMH-derived epitopes.
    The numbers above the bars denotes the number of epitopes
    obtained using an MHC ligand assay. 
  }
  \label{fig:elution}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Presentation of TMH-derived epitopes \emph{in vivo} in T cells}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure \ref{fig:t_cells_present_tmh_derived_epitopes} 
shows the percentage of TMH-derived
epitopes of the reported epitopes that for which T-cell responses were established. 
The data was obtained from the IEDB and includes only the MHC
alleles used in this study. As there are many (especially class II)
MHC alleles, only a small percentage of the full IEDB data could be used.

>>>>

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
The peptide-binding grooves of the alleles of most haplotypes
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
therefore peptide fragments derived from human transmembrane helices (TMHs) are predicted to be presented more often as would be expected 
based on their abundance in the proteome. However, the physiological reason of why membrane proteins might be over-presented is unclear. 
In this study, we show that the predicted over-presentation of TMH-derived peptides is general, as it is predicted for bacteria and viruses 
and for both MHC-I and MHC-II, and confirmed by re-analysis of epitope databases. Moreover, we show that TMHs are evolutionarily more conserved, 
because single nucleotide polymorphisms (SNPs) are present 
relatively less frequently in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytoplasmic protein regions. 
Thus, our findings suggest that 
both cytolytic and helper T cells are more tuned to respond to membrane proteins, 
because these are evolutionary more conserved. 
We speculate that TMHs are less prone to mutations 
that enable pathogens to evade T cell responses.

@@

Our immune system fights diseases and infections from pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more specific recognition of pathogens 
than the innate immune response, 
are T cells which recognize peptides, called epitopes, derived from antigenic proteins presented on Major Histocompatibility Complexes (MHC) class I and II on the cell surface.

@@

TMHs are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of, on average, 23 amino acids in length. 
TMHs can also be predicted with high accuracy from a protein sequence 
by bioinformatics approaches \cite{krogh2001predicting,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}. 
For example, a study by Jones \cite{jones2007improving} found that,
from 184 transmembrane proteins (TMPs) with known topology, 
80\% of the TMH predictions of these proteins
matched the experimental findings.
TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate that 15-39\% of all proteins 
in the human proteome contain at least one TMH \cite{ahram2006estimation}.
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble (i.e., extracellular or cytoplasmic) protein regions is unknown. 
In this study, we hypothesized that the presentation of 
TMH residues is evolutionarily preferred, 
since TMHs are less prone to undergo escape mutations. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their variability 
by the functional requirement to span a lipid bilayer. 
This limits 
many of the amino acids present in TMHs 
to have hydrophobic side chains 
\cite{hessa2007molecular,jones1994model}.
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as that will result in a dysfunctional TMH 
and render the protein inactive.

This study had two objectives. 
First, we aimed to generalize our findings by predicting
the antigenic presentation
from different kingdoms of life in both MHC-I and -II. 
From these \emph{in silico} predictions, we conclude that TMH-derived
epitopes from a human, viral and bacterial proteome
are likely to be presented more often than expected by chance 
for most alleles of MHC-I and II. 
We confirmed the presentation of TMH-derived peptides 
by re-analysis of peptides from The Immune Epitope Database (IEDB) \cite{vita2019immune}.
Second, we tested our hypothesis that TMHs 
are more evolutionary conserved than soluble protein regions.
Our analysis of human single nucleotide polymorphisms (SNPs) showed 
that random point mutations are indeed less likely
to occur within TMHs. 
These findings strengthen the emerging notion 
that TMHs are important for the T cell-mediated adaptive immune system, 
and hence are of  
overlooked importance in vaccine development.

@@

To obtain experimental confirmation that peptides stemming from TMHs 
are presented by MHC-I and MHC-II,
we mined the IEDB \cite{vita2019immune}
for confirmed human MHC-ligands.
We queried the IEDB for all linear epitopes obtained
from MHC ligand assays in healthy humans, 
carrying the MHC alleles as used in this study.
From these epitopes, we kept those that were present
exactly once in the human reference proteome
with UniProt ID UP000005640\_9606.
We predicted the topology of the protein each epitope
was found in, using \verb;TMHMM; \cite{krogh2001predicting},
from which we concluded if the epitope is overlapping with a TMH
with at least 1 amino acid.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used: 
the \emph{dbSNP} \cite{sherry2001dbsnp} database,
which contains 650 million 
cataloged non-redundant human variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}), and the databases \emph{gene} (for gene names \cite{brown2015gene})
and \emph{protein} (for proteins sequences \cite{sayers2010database}).

@@

The Immune Epitope Database (IEDB) from the National Institutes of Health contains millions of linear epitope sequences obtained
by MHC ligand assays.
For the MHC alleles used in this study, 
we obtained 54,303 and 2,484 linear epitope sequences for the MHC-I
and MHC-II alleles from human origin respectively.
There are relatively few epitopes for MHC-II, 
as MHC-II has many more different alleles than MHC-I,
whereas we selected only the human epitopes found for the 21 MHC-II alleles used in this study.

Figure (\ref{fig:2a} and \ref{fig:thm_presentation_correlation}) shows there are similar levels of
over-presentation of TMH-derived epitopes between (1) the
percentage of TMH-derived epitopes that is reported in the IEDB database
versus (2) the percentage of TMH-derived epitopes that is predicted to be presented
in MHC-I alleles.
For MHC-II alleles, there were too few epitopes per MHC allele
to result in an informative figure.

In figure \ref{fig:2b} we grouped all the epitopes 
presented by MHC-I and MHC-II alleles 
by the percentage of TMH-derived epitopes,
which are 22\% and 10\%, respectively.

@@

\subsection{Mechanism of MHC presentation of TMH-derived epitopes}
% \paragraph{TMH presentation}

@@

  \caption{
    \textbf{
      Robust prediction that TMH epitopes are presented.
    }
    The percentage of epitopes for MHC-I and -II alleles that overlap with TMHs
that are presented. The pair of horizontal red lines in each plot indicate the
lower and upper bound of the 99\% confidence interval. Note that only one line
is visible as this interval is relatively narrow. A\ref{tab:haplotype_abbreviations}).
 \textbf{(A)} Observed and predicted percentage of TMH-derived epitopes for MHC-I alleles. \textbf{(B)} MHC
ligands from IEDB corresponding to TMH-derived epitopes. The numbers above the
bars denotes the number of TMH derived epitopes obtained.

  }
  \label{fig:elution}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Presentation of TMH-derived epitopes result in T cell responses}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure \ref{fig:t_cells_present_tmh_derived_epitopes} 
shows the percentage of TMH-derived
epitopes of the reported epitopes from human origin for which T-cell responses were established. 
The data was obtained from the IEDB and includes only the MHC
alleles used in this study. As there are many (especially class II)
MHC alleles, only a small percentage of the full IEDB data could be used.
